---
title: "Exploratory Plots"
output:
  html_document:
    toc: true
    toc_depth: 5
    number_sections: true
  word_document:     
    toc: true
    toc_depth: 5
  md_document:
    variant: markdown_github
  pdf_document: 
    toc: true
    toc_depth: 5
    number_sections: true
csl: ieee.csl    
bibliography: ML.bib
---

```{r setup, include=FALSE, warning=FALSE}
#mode: selfcontained
knitr::opts_chunk$set(echo = TRUE)
```


# Setting the environment
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
load("./data/Datasets.RData")
```

# Descriptive statistics
```{r, data-loading, cache=TRUE} 
summary(training_Data)
```

# Exploratory box-plots for all the features
```{r box-plot, fig.width=15, fig.height=40}
numeric_or_not <- sapply(training_Data, is.numeric)
numeric_vars <- names(training_Data)[numeric_or_not]
featurePlot(
    x = training_Data %>% select(all_of(numeric_vars)),
    y = training_Data$classe,
    plot = "box",
    mar =c(0,0,0,0),
    ## Pass in options to bwplot()
    scales = list(y = list(relation = "free")),
    layout = c(6, round(length(numeric_vars)/5)),
    auto.key = list(columns = 2),
    par.strip.text = list(cex = 0.75)
)
```


# Density plots for the features
``````{r density-plot, fig.width=20, fig.height=40, cache=TRUE}
featurePlot(
  x = training_Data %>% select(all_of(numeric_vars)),
  y = training_Data$classe,
  plot = "density",
  ## Pass in options to xyplot() to
  ## make it prettier
  scales = list(
    x = list(relation = "free"),
    y = list(relation = "free")
  ),
  adjust = 1.5,
  pch = "|",
  layout = c(6, round(length(numeric_vars) / 5)),
  auto.key = list(columns = 5),
  par.strip.text = list(cex = 0.75)
)
```

The main project report can be found [here](index.html).

 