---
title       : _globalTimeConverter_
subtitle    : Course Project - Developing Data Products 
author      : Preethy Nair
job         : Coursera Data Science Specialization
framework   : revealjs # {io2012, html5slides, shower, dzslides, ...}
revealjs    :
  theme: night
  transition: fade
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


## globalTimeConverter
---------------------
<center>
*[globalTimeConverter](https://nairps.shinyapps.io/globaltimeconverter/): Convert times from one zone to another using Shiny*
</center>
### Course Project:
### Developing Data Products
Data Science Specialization - JHU at Coursera

<font size="10"> Author: Preethy Nair </font>

<medium> [GitHub](https://github.com/nairps) </medium>

---  

#### Main objective of this project is to build <font color="cyan">globalTimeConverter</font> that converts time from one zone to another
- Here is a world map colored by continents

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
library(RColorBrewer); library(countrycode); library(maps); library(gridExtra)
library(tidyverse); library(rworldmap); world_map <- map_data("world")
world_map$Continent <- countrycode( sourcevar = world_map$region, 
    origin = "country.name", destination = "continent" )
world_map$Continent[which(world_map$region == "Antarctica")] <- "Antarctica"
ggplot(world_map, aes(x = long, y = lat, group = group, fill = continent)) +
  geom_polygon( fill = brewer.pal(n = 7, name = "Dark2")[as.numeric(as.factor(world_map$Continent))],
    colour = "white")
```


---

- Here is [an image from Wikipedia](https://upload.wikimedia.org/wikipedia/commons/8/88/World_Time_Zones_Map.png) showing global timezones
- Each vertical segment has a different time
- Time conversions are needed when working with people across the globe


### <font color="cyan">_globalTimeConverter_</font>
---------------------------

#### <font color="cyan">_globalTimeConverter_</font> is a simple Shiny-based web application that converts time from one zone to another.

#### <font color="cyan">Organisation of the UI</font>
- **`Title panel`**: Title
- **`Navigation bar`**: Navigation tabs for the (i) main application (ii) About page and (iii) Documentation page
- **Side panel**: Location of the user input widgets: drop-down menus for `time zones` & widgets for providing date and time
- **`Main panel`**:
  - Displays HTML summary of the time conversion 
  - ***Two tab panels***  
        - 1. **A Table**:  All hourly times (N=24) for the selected date converted to the times of the selected zone
        - 2. **A Summary**: A `verbal statement` for the time conversion



---

### <font color="cyan">_globalTimeConverter_</font>
---------------------------

- All the input widgets have been initialized with a starting value

- Each session begins with the results for these initialized values displayed as output on the `main panel`.

- This application is deployed on [shinyapps.io server](https://nairps.shinyapps.io/globalTimeConverter/) and the source code is freely available under MIT license from the [GitHub]((https://github.com/nairps/data-science-specialisation-jhu/developing-data-products/globalTimeConverter) 

- A detailed documentation for the application is available on [GitHub pages](https://nairps.github.io/data-science-specialisation-jhu/developing-data-products/globalTimeConverter/).



### Applications
---------------------------
- Useful for scheduling and attending global meetings and conferences
- Helpful for people who work remotely from a different time zone    
- Convenient for facilitating collaborations across time zones

---



#### Continent-wise country stats available for time conversions
---------------------------


```{r, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
timeZones <- readRDS("data/time_zones.RDS")
df <- data.frame(Total = do.call(rbind, lapply(timeZones, length)))
df$Continent <- rownames(df); rownames(df) <- NULL
df <- df %>% mutate(Continent = fct_reorder(Continent, desc(Total))) %>% arrange(Total)
tz_table <- tableGrob(df)
tz_plot <- ggplot(df, aes(x=Continent, y=Total, fill=Continent))+
    geom_bar(stat = "identity")+ labs(y="No: of countries per continent for time conversions")+
    theme(axis.text.x = element_text(angle=45, face="bold"))+ scale_fill_brewer(palette = "Paired")
grid.arrange(tz_table,tz_plot , ncol=2, widths=c(1,2))

```

#### Thank You
---------------------------



